Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-04-24 bbbq_article.tex
Add non-standard abbreviations list, fix #204
Abreviations is an unnumbered section
Follow guidelines regarding table, progress #210

@@

This study differs in one important aspect from our previous work \cite{bianchi2017}. 
The definition of a binder differs from \cite{bianchi2017}:
in the current study, a peptide is called a binder if, for a certain haplotype, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
whereas the previous study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the results are similar, a side by side comparison was performed shown in the supplementary materials.

@@

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}.

@@

  \begin{subfigure}[t]{0.35\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc1.png}
    \label{fig:hydrophobicity_1}
  \end{subfigure}  
  \hfill
  \begin{subfigure}[t]{0.35\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc2.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \caption{ \textbf{Over-presentation of TMH-derived epitopes on most MHC-I and -II haplotypes}
    \textbf{(A)} 
    The percentage of epitopes for MHC-I and -II haplotypes that are predicted to 
    overlap with TMHs for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The red lines indicate the percentages as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \textbf{(B-C)}
    Correlation between the percentages of predicted TMH-derived epitopes
    and the hydrophobicity score of all predicted epitopes for MHC-I \textbf{(B)}
    and MHC-II haplotypes \textbf{(C)}. Red curve: linear regression analysis. Labels are shorthand for the HLA haplotypes,
    see Table \ref{tab:haplotype_abbreviations} for the names.
  }
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II haplotypes, 
for a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{M. tuberculosis}.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the exact TMH and epitope counts.

@@

To obtain experimental confirmation that peptides stemming from TMHs 
are presented in MHC-I and MHC-II,
two peptide elution studies were reanalyzed:
For MHC-I, peptides presented \emph{in vivo} 
by the (humane) haplotypes HLA-A and B 
were sequenced \cite{schellens2015comprehensive},
for MHC-II these were haplotypes DQ2.5, DQ2.2, and 
DQ7.5 \cite{bergseng2015different}.
Figure \ref{fig:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I and MHC-II elution 
studies,
for the two topology prediction tools \verb;TMHMM; \cite{krogh2001predicting} 
and \verb;PureseqTM; \cite{wang2019efficient}. 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.
See the supplementary materials (table \ref{tab:elution}) for the exact values.

@@

We addressed the question whether there is an evolutionary advantage in presenting TMHs.
We determined the conservation of TMHs 
by comparing the occurrences of SNPs located in TMHs or soluble protein regions 
for the genes coding for membrane proteins.
We obtained 911 unique gene names associated with the phrase 'membrane protein',
which are genes coding for both membrane-associated proteins (MAPs, which have no TMH) and 
transmembrane proteins (TMPs, which have at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs and 
2,237 proteins are predicted to be MAPs.
We obtained 37,630 unique variations, 
of which 9,621 are SNPs that resulted in a straightforward amino acids substitution, 
of which 6,062 were located in predicted TMPs.
See tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for the detailed numbers and distributions of SNPs.

@@

Table \ref{tab:f_tmh} shows an overview of the findings,
where a target specifies the source of the proteome,
where \verb;covid; denotes SARS-CoV-2 and \verb;myco; denotes
\emph{Mycobacterium tuberculosis}. \verb;mhc_class; denotes the MHC
class, \verb;n_spots; the number of possible 9-mers (for MHC-I) 
or 14-mers (for MHC-II) possible. \verb;n_spots_tmh; the
number of epitopes that overlapped with a TMH that were binders. 
\verb;f_tmh; the percentage of peptides that had at least 1 residue
overlapping with a TMH.

@@

Per target proteome (i.e. human, SARS-CoV-2, \emph{M tuberculosis}),
we collected all 9-mers (for MHC-I) and 14-mers (for MHC-II),
after removing the selenoproteins and proteins that are shorter
than the epitope length.
From these epitopes, per MHC haplotype,
we predicted the IC50 (in nM) using \verb;epitope-prediction; (for MHC-I)
and MHCnuggets (for MHC-II). 
Here, we show the IC50 value per haplotype that
is used to determine if a peptide binds to the haplotype's MHC
for MHC-I (see table \ref{tab:ic50_binders_mhc1}) and 
MHC-II (see table \ref{tab:ic50_binders_mhc2}).

@@

Table \ref{tab:haplotype_abbreviations} shows the
shorthand notation for the HLA haplotypes.

@@

Tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
show the exact number of binders, binders that overlap with TMHs
and the percentage of binders that overlap with TMHs, as
visualized by figure \ref{fig:bbbq_1_smart_results}.

@@

See tables \ref{tab:ncbi_counts_1} and tables \ref{tab:ncbi_counts_2}
for an overview of all amounts.
In table \ref{tab:ncbi_counts_1} there are multiple instances where
the amounts are expected to add up, yet don't, as one SNP can work on
multiple isoforms. For example, there are 9,621 unique SNPs 
found in all proteins, of which 4,219 around found in MAPs 
and 6,026 in TMPs. Apparently, 624 SNPs work on a set of isoforms that
contains both MAPs and TMPs.

@@

Table \ref{tab:snp_stats} shows the statistics for all
SNPs, where tables \ref{tab:snp_stats_per_spanner_single}
and \ref{tab:snp_stats_per_spanner_multi} show the
statistics for only single-spanners and multi-spanners respectively.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abbreviations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[h]
  \begin{tabular}{ll}
    Abbreviation & Full               \\ 
    \hline
    MAP & Membrane-associated protein \\
    TMH & Transmembrane helix         \\
    TMP & Transmembrane protein
  \end{tabular}
\end{table}

@@

This study differs in one important aspect from our previous work \cite{bianchi2017}. 
The definition of a binder differs from \cite{bianchi2017}:
in the current study, a peptide is called a binder if, for a certain haplotype, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see supplementary Tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
whereas the previous study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the results are similar, a side by side comparison was performed shown in the supplementary materials.

@@

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
Tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}.

@@

  \begin{subfigure}[t]{0.35\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc1.png}
    \label{fig:hydrophobicity_1}
  \end{subfigure}  
  \hfill
  \begin{subfigure}[t]{0.35\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc2.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \caption{ \textbf{Over-presentation of TMH-derived epitopes on most MHC-I and -II haplotypes}
    \textbf{(A)} 
    The percentage of epitopes for MHC-I and -II haplotypes that are predicted to 
    overlap with TMHs for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The red lines indicate the percentages as expected by chance.
    See supplementary Tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \textbf{(B-C)}
    Correlation between the percentages of predicted TMH-derived epitopes
    and the hydrophobicity score of all predicted epitopes for MHC-I \textbf{(B)}
    and MHC-II haplotypes \textbf{(C)}. Red curve: linear regression analysis. 
    Labels are shorthand for the HLA haplotypes,
    see the supplementary Table \ref{tab:haplotype_abbreviations} for the names.
  }
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II haplotypes, 
for a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{M. tuberculosis}.
See supplementary Table \ref{tab:tmh_binders_mhc2} 
for the exact TMH and epitope counts.

@@

To obtain experimental confirmation that peptides stemming from TMHs 
are presented in MHC-I and MHC-II,
two peptide elution studies were reanalyzed:
For MHC-I, peptides presented \emph{in vivo} 
by the (humane) haplotypes HLA-A and B 
were sequenced \cite{schellens2015comprehensive},
for MHC-II these were haplotypes DQ2.5, DQ2.2, and 
DQ7.5 \cite{bergseng2015different}.
Figure \ref{fig:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I and MHC-II elution 
studies,
for the two topology prediction tools \verb;TMHMM; \cite{krogh2001predicting} 
and \verb;PureseqTM; \cite{wang2019efficient}. 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.
See the supplementary Table \ref{tab:elution} for the exact values.

@@

We addressed the question whether there is an evolutionary advantage in presenting TMHs.
We determined the conservation of TMHs 
by comparing the occurrences of SNPs located in TMHs or soluble protein regions 
for the genes coding for membrane proteins.
We obtained 911 unique gene names associated with the phrase 'membrane protein',
which are genes coding for both membrane-associated proteins (MAPs, which have no TMH) and 
transmembrane proteins (TMPs, which have at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs and 
2,237 proteins are predicted to be MAPs.
We obtained 37,630 unique variations, 
of which 9,621 are SNPs that resulted in a straightforward amino acids substitution, 
of which 6,062 were located in predicted TMPs.
See supplementary Tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for the detailed numbers and distributions of SNPs.

@@

Supplementary Table \ref{tab:f_tmh} shows an overview of the findings,
where a target specifies the source of the proteome,
where \verb;covid; denotes SARS-CoV-2 and \verb;myco; denotes
\emph{Mycobacterium tuberculosis}. \verb;mhc_class; denotes the MHC
class, \verb;n_spots; the number of possible 9-mers (for MHC-I) 
or 14-mers (for MHC-II) possible. \verb;n_spots_tmh; the
number of epitopes that overlapped with a TMH that were binders. 
\verb;f_tmh; the percentage of peptides that had at least 1 residue
overlapping with a TMH.

@@

Per target proteome (i.e. human, SARS-CoV-2, \emph{M tuberculosis}),
we collected all 9-mers (for MHC-I) and 14-mers (for MHC-II),
after removing the selenoproteins and proteins that are shorter
than the epitope length.
From these epitopes, per MHC haplotype,
we predicted the IC50 (in nM) using \verb;epitope-prediction; (for MHC-I)
and MHCnuggets (for MHC-II). 
Here, we show the IC50 value per haplotype that
is used to determine if a peptide binds to the haplotype's MHC
for MHC-I (see supplementary Table \ref{tab:ic50_binders_mhc1}) and 
MHC-II (see supplementary Table \ref{tab:ic50_binders_mhc2}).

@@

Supplementary Table \ref{tab:haplotype_abbreviations} shows the
shorthand notation for the HLA haplotypes.

@@

Supplementary Tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
show the exact number of binders, binders that overlap with TMHs
and the percentage of binders that overlap with TMHs, as
visualized by figure \ref{fig:bbbq_1_smart_results}.

@@

See supplementary Tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}
for an overview of all amounts.
In supplementary Table \ref{tab:ncbi_counts_1} there are multiple instances where
the amounts are expected to add up, yet don't, as one SNP can work on
multiple isoforms. For example, there are 9,621 unique SNPs 
found in all proteins, of which 4,219 around found in MAPs 
and 6,026 in TMPs. Apparently, 624 SNPs work on a set of isoforms that
contains both MAPs and TMPs.

@@

Supplementary Table \ref{tab:snp_stats} shows the statistics for all
SNPs, where supplementary Tables \ref{tab:snp_stats_per_spanner_single}
and \ref{tab:snp_stats_per_spanner_multi} show the
statistics for only single-spanners and multi-spanners respectively.
